---
title: "Introduction to splicelogic"
author:
    - name: Beatriz Campillo
      orcid: 0000-0001-7323-9125
    - name: Michael I. Love
      orcid: 0000-0001-8401-0545
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: rmarkdown::html_document
vignette: |
  %\VignetteIndexEntry{Introduction to `splicelogic`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`splicelogic` is an R/Bioconductor package for detecting alternative splicing events from exon-level data stored as `GRanges` objects. Given a set of exons annotated with a coefficient column indicating differential transcript usage (DTU), `splicelogic` identifies the following types of splicing events:

-   **Skipped exons (SE)** -- exons present in one isoform but absent in another.
-   **Retained exons (RE)** -- exons present in one isoform but absent in another.
-   **Mutually exclusive exons (MXE)** -- pairs of exons where one is included at the expense of the other.
-   **Retained introns (RI)** -- intronic regions that are retained as part of an exon in an alternative isoform.
-   **Alternative 5' (A5SS)** -- exons that share the same 3' splice site but differ at the 5' splice site.
-   **Alternative 3' (A3SS)** -- exons that share the same 5' splice site but differ at the 3' splice site.

# Quick start

With DTU results attached to a *GRanges* of the exons from significant transcripts, one can use the following code to identify splice events:

```{r eval=FALSE}
exons <- preprocess_input(exons, coef_col = "estimates")
skipped <- exons |> calc_skipped_exons()
# etc.
```

# Input data

`splicelogic` assumes the user has run a differential transcript usage (DTU) or differential splicing analysis providing an error bound (e.g. adjusted p-value / FDR) and and effect estimate with direction (e.g. GLM estimated coefficient or deltaPSI) (see [upstream methods](#upstream-methods)).

It also assumes the user has obtained genomic ranges representing the exon structure of each transcript being analyzed (see [obtaining exon ranges](#obtaining-exon-ranges)).

The `exons` should be provided in a flat *GRanges* object (one range per exon), containing exon-level metadata in `mcols(exons)` such as the gene ID, transcript ID, rank in the transcript, and the estimated change or at least direction of change from the differential analysis.

Required columns for exons:

| Column | Description |
|----|----|
| `gene_id` | Gene identifier |
| `tx_id` | Transcript identifier |
| `exon_rank` | Exon rank within the transcript |
| `coef_col = "<user supplied>"` | differential effect estimate for the transcript |

The differential effect estimate column is specified by the user in `preprocess_input()`. This should be the differential effect estimate associated with the specific transcript containing the exon, and minimally could indicate the direction of effect with `+1/-1`. Positive values indicate up-regulated exons and negative values indicate down-regulated exons. All exons from the same transcript will share the same value for this column.

## Jones et al mouse long read dataset

We will use a published long read dataset and its reported splicing changes to demonstrate some of the functionality in *splicelogic*.

The citation is:

> Emma F. Jones, Timothy C. Howton, Victoria L. Flanary, Amanda D. Clark, Brittany N. Lasseigne Long-read RNA sequencing identifies region- and sex-specific C57BL/6J mouse brain mRNA isoform expression and usage **Mol Brain** 17, 40 (2024). [doi: 10.1186/s13041-024-01112-7](https://doi.org/10.1186/s13041-024-01112-7)

And information about the paper, including code and publicly available data can be found at this URL:

<https://github.com/lasseignelab/230227_EJ_MouseBrainIsoDiv>

In the abstract, Jones *et al.* describe the experiment:

> To assess differences in AS across the cerebellum, cortex, hippocampus, and striatum by sex, we generated and analyzed Oxford Nanopore Technologies (ONT) long-read RNA sequencing (lrRNA-Seq) C57BL/6J mouse brain cDNA libraries. From \> 85 million reads that passed quality control metrics, we calculated differential gene expression (DGE), differential transcript expression (DTE), and differential transcript usage (DTU) across brain regions and by sex.

## Loading data

We load the differential transcript usage analysis from the Jones *et al.* paper. Specifically, we use the cortex-specific sex comparison (F vs M) that identified the following transcripts as DTU:

<https://github.com/lasseignelab/230227_EJ_MouseBrainIsoDiv/blob/main/results/dtu_transcripts/cortex_sex_sig_features.csv>

The DTU table for this example, and the exons for the differential transcripts have been saved in the `extdata` directory of _splicelogic_.

```{r data, message=FALSE}
library(readr)
# load DTU results 
dir <- system.file("extdata", package="splicelogic")
dtu_table <- readr::read_delim(file.path(dir, "dtu_table.tsv"))
```

We load the 601 exons from GENCODE M31 annotation for the 49 differential transcripts, and attach the metadata columns.

```{r load-exons, message=FALSE}
library(plyranges)
exons_file <- "exons_M31.bed.gz"
exons_mcols_file <- "exons_mcols.tsv.gz"
exons <- plyranges::read_bed(file.path(dir, exons_file))
mcols(exons) <- DataFrame(readr::read_delim(file.path(dir, exons_mcols_file)))
```

Finally, we add `Seqinfo` for mm49, the reference mouse genome for GENCODE M31.

```{r add-seqinfo}
si <- Seqinfo::Seqinfo(genome="mm39")
seqlevels(exons) <- seqlevels(si)
seqinfo(exons) <- si
```

Next, we join the DTU results onto the `exons` GRanges. This is necessary for the _splicelogic_ functions to know which transcripts are up- or down-regulated, and to which gene the exons and transcripts belong. 

The code below matches the transcript IDs in the `exons` metadata with the DTU results table, and then binds the two tables before assigning back to `exons`.

```{r merge-dtu}

txp_idx <- match(exons$tx_id, dtu_table$tx_id)
cols_to_add <- dtu_table[txp_idx, ] |> dplyr::select(-tx_id)
merged_DF <- S4Vectors::cbind(mcols(exons), cols_to_add)
mcols(exons) <- merged_DF
```

As of Bioc 3.23 and *plyranges* version 1.32, the above code to add metadata columns from an additional table can be replaced with the following:

```{r eval=FALSE}
exons <- exons |> 
  plyranges::join_mcols_left(dtu_table, by = "tx_id")
```

_splicelogic_ is designed to take as input the exons from significantly changed transcripts, so we first filter out transcripts that were not signficant at FDR 10%.

```{r filter-sig}
exons <- exons |> filter(padj < .1)
```

# Detecting splicing events

## Preprocessing input data

The first step is to run `preprocess_input()`, which prepares the `exons` for event detection. This function checks the input data, ensures that the necessary columns are present:

```{r setup, message = FALSE}
library(splicelogic)
exons <- exons |> 
  preprocess_input(coef_col = "estimate")
```

## Individual events

Next we can run the various functions for calculating different types of splicing events. E.g. to calculate exons which are skipped in up-regulated transcripts relative to down-regulated transcripts, across all genes:

**Skipped exons**

```{r calc-se}
skipped <- exons |> calc_skipped_exons()
skipped
```

**Included exons**

```{r calc-re}
included <- exons |> calc_included_exons()
included
```

**Mutually exclusive exons**

```{r calc-mx}
mx <- exons |> calc_mx_exons()
mx
```

**Retained introns**

```{r calc-ri}
ri <- exons |> calc_retained_introns()
ri
```

**Alternative 5' and 3' splice sites**

```{r calc-a5-a3-ss}
a5ss <- exons |> calc_a5ss()
a5ss
a3ss <- exons |> calc_a3ss()
a3ss
```

## Calculate all events

```{r calc-all}
all_events <- exons |> calc_all_events()
all_events
```

```{r}
barplot(table(all_events$event))
```

# Upstream methods {#upstream-methods}

Something about what methods can be used for upstream DTU or switching analysis.

# Obtaining exon ranges with prepare_exons

`prepare_exons()` extracts exon ranges from a *TxDb* object and merges them with your DTU results table in a single call. It returns a flat *GRanges* ready for `preprocess_input()`. We demonstrate using GENCODE v32 (human genes).

The first step is to load a _TxDb_ object. Typically, a user would supply their own GTF to txdbmaker::makeTxDbFromGFF() to generate this. For this demonstration we will load a pre-constructed _TxDb_ from Bioconductor's AnnotationHub.

```{r ahub}
suppressPackageStartupMessages({
  library(AnnotationHub)
})
ah <- AnnotationHub()
txdb <- ah[["AH75191"]] # GENCODE v32 (human)
```

```{r get-txps}
suppressPackageStartupMessages({
  library(AnnotationDbi)
  library(tibble)
})
txps <- txdb |>
  AnnotationDbi::select(keys(txdb, "TXID"), c("TXNAME","GENEID"), "TXID") |>
  tibble::as_tibble() |>
  dplyr::select(tx_num = TXID, tx_id = TXNAME, gene_id = GENEID)
txps
```

```{r sim-dtu}
# simulate DTU results
dtu_table <- txps |>
  dplyr::mutate(
    padj = runif(dplyr::n()),
    effect_est = rnorm(dplyr::n())
  )
```

```{r prepare-exons}
exons <- prepare_exons(txdb, dtu_table, coef_col = "effect_est", verbose = TRUE)
exons |> preprocess_input(coef_col = "effect_est")
# followed by calculation of splicing events
```

The following section details what `prepare_exons()` does internally.

# Obtaining exon ranges manually

This section walks through the steps that `prepare_exons()` performs internally. This is useful if you need more control over the process or want to understand how exon ranges are built from a *TxDb*.

The following extracts the exons grouped by transcript from the *TxDb*:

```{r extract-exons}
suppressPackageStartupMessages({
  library(GenomicFeatures)
})
# extract exons as a GRangesList
exons_list <- GenomicFeatures::exonsBy(txdb, by="tx") # exon id, name, and rank
# dtu_table aligns with txps, which aligns with the names of the GRangesList.
# prepare_exons() handles this with alignment checks
names(exons_list) <- dtu_table$tx_id
```

Next flattening the exons:

```{r flatten-exons}
exons <- unlist(exons_list)
# swap tx_id with exon_name as the names of the GRanges
exons$tx_id <- names(exons) # store transcript ids
names(exons) <- exons$exon_name
```

Adding DTU results and gene ID:

```{r add-dtu}
txp_idx <- match(exons$tx_id, dtu_table$tx_id)
cols_to_add <- dtu_table[txp_idx,] |> dplyr::select(-c(tx_id, tx_num))
merged_DF <- cbind(mcols(exons), cols_to_add)
mcols(exons) <- merged_DF
```

# Session info

```{r session-info}
sessionInfo()
```
