---
title: "Introduction to splicelogic"
author:
    - name: Beatriz Campillo
      orcid: 0000-0001-7323-9125
    - name: Michael I. Love
      orcid: 0000-0001-8401-0545
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: rmarkdown::html_document
vignette: |
  %\VignetteIndexEntry{Introduction to `splicelogic`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`splicelogic` is an R/Bioconductor package for detecting alternative
splicing events from exon-level data stored as `GRanges` objects.
Given a set of exons annotated with a coefficient column indicating
differential transcript usage (DTU), `splicelogic` identifies
the following types of splicing events:

- **Skipped exons (SE)** -- exons present in one isoform but absent
  in another.
- **Retained exons (RE)** -- exons present in one isoform but absent
  in another.
- **Mutually exclusive exons (MXE)** -- pairs of exons where one is
  included at the expense of the other.
- **Retained introns (RI)** -- intronic regions that are retained as
  part of an exon in an alternative isoform.
- **Alternative 5' (A5SS)** -- exons that share the same 3' splice site
  but differ at the 5' splice site.
- **Alternative 3' (A3SS)** -- exons that share the same 5' splice site
  but differ at the 3' splice site.

# Quick start

With DTU results attached to a *GRanges* of the exons from significant 
transcripts, one can use the following code to identify splice events:

```{r eval=FALSE}
exons <- preprocess_input(exons, coef_col = "estimates")
skipped <- exons |> calc_skipped_exons()
# etc.
```

# Input data

`splicelogic` assumes the user has run some kind of
DTU (e.g. isoform switching) statistical analysis
providing error bound (FDR) and effect estimate
(e.g. model coefficient or deltaPSI)
(see [upstream methods](#upstream-methods)).
It also assumes the user can obtain ranges
representing the exon structure of each transcript
being analyzed
(see [obtaining exon ranges](#obtaining-exon-ranges).
The exons should be in a flat *GRanges* object
(one range per exon), containing exon-level metadata
such as the gene ID, transcript ID, and rank in the
transcript.
Information from the DTU analysis should also be present in the metadata.
Code below demonstrates generating this from a GTF file and attaching metdata
from a DTU results table.

| Column      | Description                                        |
|-------------|----------------------------------------------------|
| `gene_id`   | Gene identifier                                    |
| `tx_id`     | Transcript identifier                              |
| `exon_rank` | Exon rank within the transcript                    |
| *coef_col*  | DTU effect estimate (column name is user-defined)  |

<!--
considering if exon_rank should be inputed by user or calculated internally
currently, it is required as input
-->

The coefficient input in *coef_col* indicates the
differential transcript usage (DTU) of the specific
transcript containing the exon. All exons from the
same transcript share the same coefficient value.
This values comes from a prior differential
transcript usage
analysis, such as that performed by 
[satuRn](https://bioconductor.org/packages/satuRn).
Positive coefficient values indicate upregulated exons and negative
values indicate downregulated exons. 

## Jones et al mouse long read dataset

We will use a published long read dataset and it's reported splicing 
changes to demonstrate some of the functionality in *splicelogic*.

The citation is:

> Emma F. Jones, Timothy C. Howton, Victoria L. Flanary, 
> Amanda D. Clark, Brittany N. Lasseigne
> Long-read RNA sequencing identifies region- and sex-specific 
> C57BL/6J mouse brain mRNA isoform expression and usage
> **Mol Brain** 17, 40 (2024). 
> [doi: 10.1186/s13041-024-01112-7](https://doi.org/10.1186/s13041-024-01112-7)

And information about the paper, including code and publicly
available data can be found at this URL:

<https://github.com/lasseignelab/230227_EJ_MouseBrainIsoDiv>

In the abstract, Jones *et al.* describe the experiment:

> To assess differences in AS across the cerebellum, cortex, hippocampus, 
> and striatum by sex, we generated and analyzed Oxford Nanopore Technologies 
> (ONT) long-read RNA sequencing (lrRNA-Seq)
> C57BL/6J mouse brain cDNA libraries.
> From > 85 million reads that passed quality control metrics, we calculated 
> differential gene expression (DGE), differential transcript expression (DTE), 
> and differential transcript usage (DTU) across brain regions and by sex.

## Loading data
```{r data, message = FALSE}
library(readr)
# load DTU results 
dir <- system.file("extdata", package="splicelogic")
dtu_table <- readr::read_delim(file.path(dir, "dtu_table.tsv"))

library(plyranges)
# load exon ranges and metadata, and set seqinfo
exons <- read_bed(file.path(dir, "exons_M31.bed.gz"))
mcols(exons) <- DataFrame(
  readr::read_delim(file.path(dir, "exons_mcols.tsv.gz"))
)
# M31 is paired with GRCm39 / mm39
si <- Seqinfo::Seqinfo(genome="mm39")
seqlevels(exons) <- seqlevels(si)
seqinfo(exons) <- si
```

Next, we join the DTU results into the exon GRanges metadata. 
This is necessary for the splicelogic functions to know which transcripts 
are up or downregulated, and to which gene they belong. The code below matches 
the transcript IDs in the exon metadata with those in the DTU results table, 
and adds the relevant columns from the DTU table to the exon metadata.

```{r merge} 
# insert dtu results into exon metadata
txp_idx <- match(exons$tx_id, dtu_table$tx_id)
add_columns <- dtu_table[txp_idx, ] |>
  dplyr::select(-c(tx_id))
merged_DF <- cbind(mcols(exons), add_columns)
mcols(exons) <- merged_DF
```

As of Bioc 3.23 and _plyranges_ version 1.32, the above can be replaced with:

```{r eval=FALSE}
exons <- exons |> 
  plyranges::join_mcols_left(dtu_table, by = "tx_id")
```

# How to start detecting splicing events

`splicelogic` works on exons from significantly changed transcripts, 
so we first filter out transcripts that were not signficant:

```{r filter-sig}
exons <- exons |> filter(padj < .05)
```

## Preprocessing input data
The first step is to run `preprocess_input()`, which prepares the exon data for
event detection. This function checks the input data, ensures that the necessary
```{r setup, message = FALSE}
library(splicelogic)
processed_exons <- preprocess_input(exons, coef_col = "estimate")
```

## Calculate skipped exons
After preprocessing, you can run the various
functions to calculate different types of
splicing events.
To calculate skipped exons:
```{r calc-se}
skipped_exons <- calc_skipped_exons(processed_exons)
skipped_exons
```

## calculate included exons
```{r calc-re}
included_exons <- calc_included_exons(processed_exons)
included_exons
```

## Calculate mutually exclusive exons
```{r calc-mx}
mx_exons <- calc_mx_exons(processed_exons)
mx_exons
```

## Calculate retained introns
```{r calc-ri}
retained_introns <- calc_retained_introns(processed_exons)
retained_introns
```

## Calculate alternative 5' splice sites
```{r calc-a5ss}
a5ss <- calc_a5ss(processed_exons)
a5ss
```

## Calculate alternative 3' splice sites
```{r calc-a3ss}
a3ss <- calc_a3ss(processed_exons)
a3ss
```

## Calculate all events
```{r calc-all}
all_events <- calc_all_events(processed_exons)
all_events
```

# Upstream methods

Something about what methods can be used for upstream DTU or switching analysis.



# Obtaining exon ranges with prepare_exons

`prepare_exons()` extracts exon ranges from a *TxDb* object and merges
them with your DTU results table in a single call. It returns a flat
*GRanges* ready for `preprocess_input()`.

```{r prepare-exons}
library(AnnotationHub)
ah <- AnnotationHub()
txdb <- ah[["AH75191"]]
# extract transcript table as tibble
txps <- txdb |>
  AnnotationDbi::select(keys(txdb, "TXID"), c("TXNAME","GENEID"), "TXID") |>
  tibble::as_tibble() |>
  dplyr::select(tx_num = TXID, tx_id = TXNAME, gene_id = GENEID)
# simulate DTU results
dtu_table <- txps |>
  dplyr::mutate(
    padj = runif(dplyr::n()),
    effect_est = rnorm(dplyr::n())
  )

exons <- prepare_exons(txdb, dtu_table, coef_col = "effect_est", verbose = TRUE)
gr <- preprocess_input(exons, coef_col = "effect_est")
```

For a step-by-step breakdown of what `prepare_exons()` does internally,
see [Obtaining exon ranges manually](#obtaining-exon-ranges-manually).

# Obtaining exon ranges manually {#obtaining-exon-ranges-manually}

This section walks through the steps that `prepare_exons()` performs
internally. This is useful if you need more control over the process
or want to understand how exon ranges are built from a *TxDb*.

```{r load-txdb}
# first we get an example TxDB for our exon ranges
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(GenomicFeatures)
  library(tibble)
})
# here we will use the transcript database for GENCODE v32 from AHub
# typically, you should supply your own GTF to makeTxDbFromGFF()
ah <- AnnotationHub()
txdb <- ah[["AH75191"]]
```

The following extracts the exons from the *TxDb*:

```{r extract-exons}
# extract exons as a GRangesList
ebt <- GenomicFeatures::exonsBy(txdb, by="tx") # exon id, name, and rank
```

Here we build a transcript table, likely this is already
provided by an upstream DTU method.

```{r build-txp-table}
# extract transcript table as tibble
txps <- txdb |>
  AnnotationDbi::select(keys(txdb, "TXID"), c("TXNAME","GENEID"), "TXID") |>
  tibble::as_tibble() |>
  dplyr::select(tx_num = TXID, tx_id = TXNAME, gene_id = GENEID)
```

Suppose we already have DTU results for transcripts:

```{r sim-dtu}
# here just simulated results
dtu_table <- txps |>
  dplyr::mutate(
    padj = runif(dplyr::n()),
    effect_est = rnorm(dplyr::n())
  )
```

```{r check-names}
# check the concordance...
all.equal(dtu_table$tx_num, as.integer(names(ebt)))
# and if aligned, set new names for exons
exons <- ebt
names(exons) <- dtu_table$tx_id
```

Next flattening the exons:

```{r flatten-exons}
exons <- unlist(exons)
exons$tx_id <- names(exons)
names(exons) <- exons$exon_name
```

Adding DTU results and gene ID:

```{r add-dtu}
# arrange `dtu_table` as in `exons`, including duplicates
txp_idx <- match(exons$tx_id, dtu_table$tx_id)
add_columns <- dtu_table[txp_idx,] |>
  dplyr::select(-c(tx_id, tx_num))
merged_DF <- cbind(mcols(exons), add_columns)
mcols(exons) <- merged_DF
```

# Session info

```{r session-info}
sessionInfo()
```
