---
title: "Introduction to `splicelogic`"
author:
    - name: Beatriz Campillo
      orcid: 0000-0001-7323-9125
    - name: Michael I. Love
      orcid: 0000-0001-8401-0545
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: rmarkdown::html_document
vignette: |
  %\VignetteIndexEntry{Introduction to `splicelogic`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`splicelogic` is an R/Bioconductor package for detecting alternative
splicing events from exon-level data stored as `GRanges` objects.
Given a set of exons annotated with a coefficient column indicating
differential transcript usage (DTU), `splicelogic` identifies
the following types of splicing events:

- **Skipped exons (SE)** -- exons present in one isoform but absent
  in another.
- **Mutually exclusive exons (MXE)** -- pairs of exons where one is
  included at the expense of the other.
- **Retained introns (RI)** -- intronic regions that are retained as
  part of an exon in an alternative isoform.
- **Alternative 3' (A3SS)** -- exons that share the same 5' splice site
  but differ at the 3' end.
- **Alternative 5' (A5SS)** -- exons that share the same 3' splice site
  but differ at the 5' end.

# Input data

`splicelogic` assumes the user has run some kind of DTU (e.g. isoform switching) 
statistical analysis providing error bound (FDR) and 
effect estimate (e.g. model coefficient or deltaPSI)
(see [upstream methods](#upstream-methods)). 
It also assumes the user can obtain ranges representing the exon structure of each 
transcript being analyzed (see [obtaining exon ranges](#obtaining-exon-ranges).
The exons should be in a flat *GRanges* object (one range per exon), containing
exon-level metadata such as the gene ID, transcript ID, and rank in the transcript. 
Information from the DTU analysis should also be present in the metadata.
Code below demonstrates generating this from a GTF file and attaching metdata
from a DTU results table.

| Column      | Description                                        |
|-------------|----------------------------------------------------|
| `gene_id`   | Gene identifier                                    |
| `tx_id`     | Transcript identifier                              |
| `exon_rank` | Exon rank within the transcript                     |
| *coef_col*  | DTU effect estimate (column name is user-defined) 

<!--
considering if exon_rank should be inputed by user or calculated internally
currently, it is required as input
-->

The coefficient input in *coef_col* indicates the differential transcript usage (DTU) of the specific
transcript containing the exon. All exons from the same transcript share the
same coefficient value. This values comes from a prior differential transcript usage
analysis, such as that performed by 
[satuRn](https://bioconductor.org/packages/satuRn).
Positive coefficient values indicate upregulated exons and negative
values indicate downregulated exons. 

## Obtaining exon ranges

Here we demontrate reading in exon ranges from a GTF and adding example DTU results.

```{r}
# look up annotations of exons and transcripts
library(AnnotationHub)
ah <- AnnotationHub()
query(ah, c("GENCODE", "Homo sapiens", "GTF", "v32"))
# download an example TxDB for generating exon ranges
txdb <- ah[["AH75191"]]
library(GenomicFeatures)
ebt <- GenomicFeatures::exonsBy(txdb, by="tx") # exon id, name, and rank
```

```{r}
library(tibble)
txps <- txdb |>
  AnnotationDbi::select(
    keys(txdb, "TXID"), 
    c("TXNAME","GENEID"), 
    "TXID"
    ) |>
  tibble::as_tibble() |>
  dplyr::select(
    tx_number = TXID,
    tx_id = TXNAME,
    gene_id = GENEID
  )
```

Suppose DTU results:

```{r}
# here just simulated results
txps <- txps |>
  dplyr::mutate(
    padj = runif(dplyr::n()),
    effect_est = rnorm(dplyr::n())
  )
```

```{r}
all.equal(txps$tx_number, as.numeric(names(ebt)))
names(ebt) <- txps$tx_id
```

Next flattening the exons:

```{r}
exons <- unlist(ebt)
exons$tx_id <- names(exons)
names(exons) <- exons$exon_name
exons[1:3] # show the top 3
```

Adding DTU results and gene ID:

```{r}
add_columns <- txps[match(exons$tx_id, txps$tx_id),]
merged_DF <- cbind(mcols(exons), add_columns)
mcols(exons) <- merged_DF
```

## Upstream methods

Something about what methods can be used for upstream DTU or switching analysis.

# How to start detecting splicing events

```{r setup, message = FALSE}
library(splicelogic)
```

## Preprocessing
We can create mock data for demonstration purposes using the
`create_mock_data()` function. This function generates a `GRanges` object with the required 
metadata columns and random coefficient values for each transcript, making sure that at least 
one transcript is upregulated (coef>0) and one is downregulated (coef<0) for each gene.

```{r preprocess, warning=FALSE}
gr <- create_mock_data( n_genes = 2, n_tx = 4, n_exons = 6 )

mcols(gr)
```

## Skipped exons

`calc_skipped_exons()` identifies exons that are present in
downregulated transcripts but absent (skipped) in upregulated ones.

Using the gr object generated by `create_mock_data()`,
we can generate skiped exon events with the `generate_skipped_exons()` function, 
which modifies the input `GRanges` object to include skipped exon events.
 Then we can run the detection function `calc_skipped_exons()` on the modified `GRanges` object:
```{r skipped-exons}
gr_se <- generate_skipped_exons(gr, n_se = 2)


gr_se <- preprocess_input(gr_se, coef_col = "coefs")

se_result <- calc_skipped_exons(gr_se, coef_col = "coefs")
se_result
```

The result is a `GRanges` object containing only the exons flagged as
skipped, with an `event` column set to `"skipped_exon"`. The `"tx_id"` refers to the transcript 
that includes the skipped exon, `"tx_event"` indicates the transcripts in respect to which the exon 
is skipped (i.e. the downregulated transcripts that do not include the exon).

```{r skipped-exons-check, eval=FALSE}
mcols(se_result)[, c("gene_id", "tx_id", "n_txp", "event", "tx_event")]
```

## Mutually exclusive exons

`calc_mutually_exclusive()` detects pairs of exons where one exon is
included in one isoform and a different exon occupies the equivalent
position in another isoform.

The `mx_mock_data()` dataset is designed to contain mutually exclusive
exon events:

```{r mx-exons, eval=FALSE}
gr_mx <- mx_mock_data()
gr_mx <- preprocess_input(gr_mx, coef_col = "coefs")
mx_result <- calc_mutually_exclusive(gr_mx, coef_col = "coefs")
mx_result
```

```{r mx-exons-check, eval=FALSE}
mcols(mx_result)[, c("gene_id", "tx_id", "exon_rank", "event")]
```

## Retained introns

`calc_retained_introns()` finds exons in upregulated transcripts that
overlap intronic regions of downregulated transcripts, indicating
intron retention.

```{r retained-introns, eval=FALSE}
gr_ri <- create_mock_data(3, 6, 3)
gr_ri <- preprocess_input(gr_ri, coef_col = "coefs")
gr_ri <- generate_retained_introns(gr_ri, n_ri = 3)
ri_result <- calc_retained_introns(gr_ri, coef_col = "coefs")
ri_result
```

```{r retained-introns-check, eval=FALSE}
if (length(ri_result) > 0) {
    mcols(ri_result)[, c("gene_id", "tx_id", "exon_rank", "event")]
}
```

## Alternative 3' and 5' splice sites

**NOTE: I think it's more common to list 5 then 3**

`calc_a3ss_a5ss()` detects exons that share one boundary (start or
end) with an exon in another isoform but differ at the opposite end,
indicating alternative splice site usage.

```{r a3ss-a5ss, eval=FALSE}
gr_ss <- create_mock_data(3, 3, 6)
gr_ss <- preprocess_input(gr_ss, coef_col = "coefs")
gr_ss <- generate_a3ss(gr_ss, n_a3ss = 3)
ss_result <- calc_a3ss_a5ss(gr_ss, coef_col = "coefs")
ss_result
```

```{r a3ss-a5ss-check, eval=FALSE}
if (length(ss_result) > 0) {
    mcols(ss_result)[, c("gene_id", "tx_id", "exon_rank", "event")]
}
```

# Handling cases with no events

When the input data does not contain any splicing events, each
detection function returns an empty `GRanges` object:

```{r no-events, eval=FALSE}
gr_none <- no_event_mock_data()
gr_none <- preprocess_input(gr_none, coef_col = "coefs")

calc_skipped_exons(gr_none, coef_col = "coefs")
calc_mutually_exclusive(gr_none, coef_col = "coefs")
```

# Session info

```{r session-info}
sessionInfo()
```
