---
title: "splicelogic: transcript sets to splice events"
author:
    - name: Beatriz Campillo
      orcid: 0000-0001-7323-9125
    - name: Michael I. Love
      orcid: 0000-0001-8401-0545
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: rmarkdown::html_document
vignette: |
  %\VignetteIndexEntry{Introduction to `splicelogic`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

*splicelogic* turns sets of transcripts into discrete splicing events.
Unlike event-based tools that work at the junction level, *splicelogic*
operates on whole transcript structures: within each gene it compares
two groups of transcripts and all of their exons, so events are derived
with full isoform context. It detects skipped exons, included exons,
mutually exclusive exons, retained introns, alternative 5' and 3'
splice sites, and alternative transcription start and end sites.

*splicelogic* allows the two groups of transcripts to be compared to be
defined upstream in one of two ways. They can come from an explicit
partition of the transcripts into two sets, e.g. a set of novel transcripts
against a set of annotated reference transcripts, or transcripts found in
disease samples relative to those found in control samples. Or they can come
from a differential transcript usage (DTU) analysis, which gives each
transcript an effect estimate whose sign defines the comparison. Because it
takes transcript-level results tables as input, *splicelogic* is compatible
with any upstream DTU method (including *DRIMSeq*, *DEXSeq*, *satuRn*, and
*edgeR*), supporting flexible experimental designs.

*splicelogic* operates on exon-level data stored as *GRanges* objects.
Given a set of exons carrying the gene ID, transcript ID, exon rank,
and a value indicating the direction of change, *splicelogic* identifies
the following types of splicing events:

-   **Skipped exons (SE)** -- exons skipped in up-regulated transcripts
-   **Included exons (IE)** -- exons included in up-regulated transcripts
-   **Mutually exclusive exons (MXE)** -- pairs of exons where one is
    included and another is excluded in up-regulated transcripts
    compared to down-regulated transcripts
-   **Retained introns (RI)** -- intronic regions that are retained
    as part of an exon in up-regulated transcripts
-   **Alternative 5' (A5SS)** -- exons in up-regulated transcripts
    that share the same 3' splice site but differ at the 5' splice
    site from exons in down-regulated transcripts
-   **Alternative 3' (A3SS)** -- as above, but differing in the 3'
    splice site

These are detected using a series of functions, e.g. `find_se()`,
`find_ie()`, etc., summarized in
[the `find_*()` functions](#find-functions).

Several alternative tools exist for detecting splicing events from
RNA-seq data and analyzing their consequences:

-   [IsoformSwitchAnalyzeR](https://www.bioconductor.org/packages/IsoformSwitchAnalyzeR)
    — a comprehensive workflow that performs DTU testing, annotates
    the resulting isoform switches with splicing event types, and
    predicts functional consequences such as domain loss and NMD.

-   [GeneStructureTools](https://www.bioconductor.org/packages/GeneStructureTools)
    — takes differential splicing results from read-based tools such
    as Whippet or leafcutter (which detect splicing events from
    junction counts), classifies the event types, and assesses their
    structural and functional consequences such as ORF changes, NMD
    potential, and UTR structure.

-   [isoformic](https://github.com/luciorq/isoformic)
    — a visualization and functional interpretation pipeline for
    pre-computed differential transcript expression results. It
    organizes transcripts by biotype (protein-coding, lncRNA, NMD,
    etc.) and provides expression profile plots and functional
    enrichment analysis.

*splicelogic* differs from these in its main input and focus: it does not perform
DTU testing, and also does not take pre-computed splicing events as input.
Instead, it works from two groups of transcripts (either an explicit partition
or transcript-level DTU results already mapped onto exon structures) and
directly classifies the type of splicing event from the comparison of the two
groups.

By operating on exon-level *GRanges* with attached DTU statistics, 
*splicelogic* provides a flexible framework for users to identify 
and interpret splicing events in the context of their specific 
experimental design and choice of DTU method. 
*GRanges* are a common data structure in Bioconductor for representing 
genomic features such as exons and introns, making *splicelogic*
compatible with a wide range of already available tools and 
existing workflows. For example, it can be combined with *plyranges* 
for dplyr-like operations on exons/introns following identification 
of regulated exon and introns, and *Biostrings* for extraction of 
sequence surrounding splice events.

# Quick start

*splicelogic* turns a set of transcripts into discrete splice events.
There are two entry points, depending on what is available upstream.
Both build a flat *GRanges* of exons that `preprocess()` prepares for
the `find_*()` functions, and from that point on the workflow is
identical.

## Finding events from transcript sets

To find splicing events from two sets of transcripts we use
`prepare_exons_by_partition()`. `down` is the anchor, or reference
group, and `up` holds the alternative transcripts that are read
against it: the function assigns `estimate = -1` to `down` and
`estimate = +1` to `up`. Each set is given either as a *GRanges* of 
exons or as a character vector of transcript IDs (the latter needs a
*TxDb* to look up the exon coordinates):

```{r quick-start-sets}
#| eval: false
exons <- prepare_exons_by_partition(
  up = <GRanges OF EXONS, OR TX IDs>,
  down = <GRanges OF EXONS, OR TX IDs>
) |>
  preprocess(coef_col = "estimate")

# find e.g. skipped exons:
skipped <- exons |> find_se()
```

This route is worked through in
[finding events from transcript sets](#transcript-sets).

## Finding events from DTU results

Each transcript carries a numeric effect estimate from a differential
transcript usage analysis, and the sign of that estimate defines the
comparison:

```{r quick-start-dtu}
#| eval: false
exons <- prepare_exons(
  txdb = <A TxDb OBJECT>,
  dtu_table = <DTU_TABLE>,
  coef_col = "estimate"
)

exons <- preprocess(exons, coef_col = "estimate")

# find e.g. skipped exons:
skipped <- exons |> find_se()
```

This route is worked through in full with a published dataset in
[Finding events after DTU analysis](#dtu-analysis).


# Input data

Within each gene, *splicelogic* compares two groups of transcripts
and works out the splice events that distinguish them. It does not
decide what those two groups are; that is settled upstream, in one of
two ways:

1.  By an explicit **partition of the transcripts** into two sets,
    with no differential analysis behind it. Direction comes from set
    membership alone, and `prepare_exons_by_partition()` fills in the
    estimates.

2.  By a **differential transcript usage (DTU)** or differential
    splicing analysis, which gives each transcript an effect estimate
    with a direction — a GLM coefficient, a change in
    percent-spliced-in ($\Delta$PSI), and so on — usually alongside a
    statistic such as an adjusted p-value for restricting to a
    significant set. See [upstream DTU methods](#upstream-methods).

*splicelogic* also assumes the user has information about the
genomic ranges for the exons of each transcript. _splicelogic_
provides a set of helper functions for generating these exon ranges,
see [obtaining exon ranges](#obtaining-exon-ranges).

Either way, the `exons` are provided in a flat *GRanges* object (one
range per exon), containing exon-level metadata in `mcols(exons)`:
the gene ID, transcript ID, rank in the transcript, and a value
giving the direction of change for the transcript.

Required columns for exons:

| Column                         | Description                                     |
|------------------------------------|------------------------------------|
| `gene_id`                      | Gene identifier                                 |
| `tx_id`                        | Transcript identifier                           |
| `exon_rank`                    | Exon rank within the transcript                 |
| `coef_col = "<user supplied>"` | Effect estimate, or direction of change, for the transcript |

The effect estimate column is named by the user in `preprocess()`.
This should be the value associated with the specific transcript
containing the exon, and minimally could indicate the direction of
effect with `+1/-1` — which is exactly what
`prepare_exons_by_partition()` supplies when there is no differential
analysis to draw on. Positive values indicate up-regulated exons and
negative values indicate down-regulated exons; the negative group is
the reference, whose exons are tested against those of the positive
group. All exons from the same transcript will share the same value
for this column.

# The `find_*()` functions {#find-functions}

One function detects each event type. All of them take as their
first argument an exon *GRanges* that has been through
`preprocess()`, and return event *GRanges* in the same shape (see
[output format](#output-format)), so they can be called in any order
or all at once with `find_all_events()`.

Throughout, *up* and *down* refer to the sign of the effect
estimate: down transcripts are the reference group, and the events
are what the up transcripts do relative to them.

| Function | `event_type` | Detects | Exons returned |
|-----------|-----------|-----------------------|-----------------------|
| `find_se()` | `"se"` | An exon of a down transcript that is absent from an up transcript of the same gene, whose two flanking exons are adjacent in that up transcript | The skipped exon, from the down transcript |
| `find_ie()` | `"ie"` | The mirror image of SE: an exon of an up transcript absent from a down transcript of the same gene | The included exon, from the up transcript |
| `find_mxe()` | `"mxe"` | Two non-overlapping exons, one in each transcript, sitting between the same pair of flanking exons | Both exons of the pair, one row each, interleaved |
| `find_ri()` | `"ri"` | An intron of a down transcript that falls entirely within a single exon of an up transcript of the same gene | The exon retaining the intron, from the up transcript |
| `find_a5ss()` | `"a5ss"` | An up exon whose 5' splice site (donor) differs from that of an overlapping down exon of the same gene. A last exon has no donor, so it is never reported here | The exon with the alternative donor, from the up transcript |
| `find_a3ss()` | `"a3ss"` | The reverse: a differing 3' splice site (acceptor). A first exon has no acceptor, so it is never reported here | The exon with the alternative acceptor, from the up transcript |
| `find_atss()` | `"atss"` | The first exon of an up transcript beginning transcription at a different coordinate from the first exon of a down transcript of the same gene. No overlap is required, so an alternative promoter elsewhere in the gene is still found | The first exon, from the up transcript |
| `find_ates()` | `"ates"` | The mirror image: a last exon ending transcription at a different coordinate | The last exon, from the up transcript |
| `find_all_events()` | all of the above | Runs the eight finders in turn and binds the results | All exons returned by the individual finders |

Neither boundary of an exon has to match its partner, so an exon that
has moved both its acceptor and its donor is reported by `find_a5ss()`
*and* by `find_a3ss()`. Because `exon_rank` runs in transcription
order, the first/last restrictions above hold on either strand: on the
minus strand the first exon is the rightmost one, and its transcription
start is its genomic end.

`find_se()`, `find_ie()`, `find_mxe()` and `find_all_events()` also
take a `type` argument, controlling how an exon flanking a candidate
is matched to an exon of the partner transcript: `"boundary"` (the
default) requires only the shared splice site to coincide, `"over"`
accepts any overlap, and `"in"` requires the two exons to be
identical.

Every finder has a long-form alias that spells the event type out —
`find_skipped_exons()`, `find_included_exons()`,
`find_mutually_exclusive_exons()`, `find_retained_introns()`,
`find_alternative_5_prime_splice_sites()`,
`find_alternative_3_prime_splice_sites()`,
`find_alternative_start_sites()` and `find_alternative_end_sites()` —
for use where a script
reads better with the full name. They are the same functions, so
`find_se()` and `find_skipped_exons()` are interchangeable.

Worked examples of all of these are in
[finding individual events](#individual-events).

# Output format {#output-format}

All `find_*()` functions return a *GRanges* whose ranges are the
exons detected to be involved in an event. Each row therefore carries the user's
original exon-level metadata — including `tx_id`, the transcript
that the detected exon belongs to — together with the following
event-specific columns added in `mcols()` that describe the
*paired* transcript against which the event is defined:

| Column            | Description                                                                                                                                                                  |
|-------------------|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| `event_type`      | Type of splicing event detected (`"se"`, `"ie"`, `"mxe"`, `"ri"`, `"a5ss"`, or `"a3ss"`).                                                                                    |
| `event_tx_id`     | Transcript ID of the *paired* transcript that, together with the detected exon's transcript (`tx_id`), defines the event.                                          |
| `event_estimate`  | DTU coefficient (`estimate`) of the paired transcript.                                                                                                                       |
| `event_<col>`     | One column per name in `metadata(gr)$additional_columns`, prefixed with `event_`, carrying the corresponding value from the paired transcript.                               |

In short: the original metadata columns describe the transcript
that *contains* the returned exon, while all columns with the prefix
`event_*` describe the paired transcript against which the event is defined.

# Finding splicing events

The rest of the vignette works through two examples that reach the
same `find_*()` functions by different routes. The first starts from
two sets of transcripts, partitioned into `up` and `down` by the user. 
The second starts from a DTU analysis, where every transcript carries a numeric
effect estimate. Both produce the same kind of exon-level *GRanges*, and from
`preprocess()` onwards the workflow is identical.

## Finding events from transcript sets {#transcript-sets}
 
Often what we have is two sets of transcripts we want to compare, e.g. a set
of novel transcripts against a set of annotated reference transcripts, or
transcripts found in disease samples relative to those found in control samples.
We can use `prepare_exons_by_partition()` to build the input for *splicelogic*.

### Input format

The exons are as described in [input data](#input-data), except that
no coefficient column is needed: `prepare_exons_by_partition()` takes
the two sets as `up` and `down`, assigns `estimate = +1` and
`estimate = -1` respectively, and returns a single *GRanges* ready for
`preprocess()` with `coef_col = "estimate"`.

Which set goes where matters. `down` is the anchor, or reference
group, and `up` the alternative transcripts read against it: the
finders take the exons of the `down` transcripts and ask which of them
are missing from, or truncated in, the `up` transcripts of the same
gene.

The simplest way to use `prepare_exons_by_partition()` is to pass the two sets
as character vectors of transcript IDs, along with a *TxDb* object that
*splicelogic* uses to look up the exon coordinates. A *TxDb* can be built from
a GTF file (e.g. from Ensembl or GENCODE) using:

```{r make-txdb}
#| eval: false
txdb <- txdbmaker::makeTxDbFromGFF("path/to/annotation.gtf")
```

Alternatively, if exon coordinates are already available as a *GRanges*, those
can be passed directly. 

For more details on preparing exons as input see the vignette section 
[comparing two_sets](#comparing-two-sets).

### A synthetic exon set

To demonstrate this route we use a small synthetic exon set saved 
as a small BED file within the package, `mock_exons.bed`.
It contains a single gene, `gene_1`, on the `+` strand,
with five transcripts all derived from the same five-exon baseline
(exons of 100 bp separated by introns of 100 bp):

-   `tx_0` — the anchor transcript, with all five exons present
-   `tx_1` — exon 3 skipped
-   `tx_2` — exon 3 with an alternative 5' splice site
-   `tx_4` — exon 4 with an alternative 3' splice site
-   `tx_5` — exon 3 skipped *and* exon 4 with an alternative 3' splice site

The script that generated the files is `inst/scripts/make-mock-data.R`.
Because BED can only carry chrom/start/end/name/score/strand, the
ranges and the metadata columns *splicelogic* needs are stored in two
files and joined on read — the same pairing used for the Jones exons
below.

```{r mock-load}
#| message: false
library(splicelogic)
library(plyranges)
dir <- system.file("extdata", package = "splicelogic")
mock_exons <- plyranges::read_bed(file.path(dir, "mock_exons.bed.gz"))
GenomicRanges::mcols(mock_exons) <- S4Vectors::DataFrame(
  readr::read_tsv(
    file.path(dir, "mock_exons_mcols.tsv.gz"),
    show_col_types = FALSE
  )
)
mock_exons |>
  dplyr::select(-gene_id)
```
We can visualize the five transcripts with `wiggleplotr::plotTranscripts()`.

```{r mock-transcripts}
#| warning: false
#| fig.height: 3
#| fig.alt: 'Exon structures of the five synthetic transcripts of gene_1'
mock_grl <- S4Vectors::split(mock_exons, mock_exons$tx_id)
wiggleplotr::plotTranscripts(mock_grl, rescale_introns = FALSE)
```

### Building the input by partition

`tx_0` is the baseline that every other transcript was derived from,
which makes it the anchor: we put it on its own in `down` and
all of the remaining transcripts in `up`, so that each variant is read
against the baseline.

```{r mock-partition}
mock <- prepare_exons_by_partition(
  up = mock_exons |> dplyr::filter(tx_id != "tx_0"),
  down = mock_exons |> dplyr::filter(tx_id == "tx_0")
) |>
  preprocess(coef_col = "estimate")
mock |> head(2)
```

### Finding events

`find_all_events()` wraps the eight event finders
([the `find_*()` functions](#find-functions)), which can also be
called one at a time — see
[finding individual events](#individual-events) below:

```{r mock-events}
mock_events <- mock |> find_all_events()
mock_events
```

Every event built into the set is recovered. Exon 3 of `tx_0` is
reported as a skipped exon twice, once against `tx_1` and once against
`tx_5`, because both of those transcripts skip it. The shortened exon
3 of `tx_2` is reported as an alternative 5' splice site, and the
shortened exon 4 as an alternative 3' splice site in both `tx_4` and
`tx_5`. As explained in [output format](#output-format), the skipped
exon rows belong to `tx_0`, the down transcript that actually contains
the exon, while the alternative splice site rows belong to the up
transcripts; in both cases the partner transcript is in `event_tx_id`.

Notice that the same exon appears on more than one row, once per partner
transcript, and that a single transcript pair can carry more than one
event. Both are expected, and
[collapsing redundant events](#collapsing-events) and
[identifying complex events](#complex-events) below show how to
summarise each.

## Finding events after DTU analysis {#dtu-analysis}

### Jones et al mouse long read dataset {#jones}

The second route starts from a real DTU analysis, where the direction
of the comparison comes from the effect estimates rather than from a
partition we supply. For this we use a published mouse long read
dataset and its reported splicing changes.

The citation is:

> Emma F. Jones, Timothy C. Howton, Victoria L. Flanary, Amanda D. Clark, Brittany N. Lasseigne Long-read RNA sequencing identifies region- and sex-specific C57BL/6J mouse brain mRNA isoform expression and usage **Mol Brain** 17, 40 (2024). [doi: 10.1186/s13041-024-01112-7](https://doi.org/10.1186/s13041-024-01112-7)

Information about the paper, including code and publicly available
data can be found at this URL:

<https://github.com/lasseignelab/230227_EJ_MouseBrainIsoDiv>

In the abstract, Jones *et al.* describe the experiment:

> To assess differences in AS across the cerebellum, cortex, hippocampus, and striatum by sex, we generated and analyzed Oxford Nanopore Technologies (ONT) long-read RNA sequencing (lrRNA-Seq) C57BL/6J mouse brain cDNA libraries. From \> 85 million reads that passed quality control metrics, we calculated differential gene expression (DGE), differential transcript expression (DTE), and differential transcript usage (DTU) across brain regions and by sex.

Here we will focus on the cortex-specific sex comparison, comparing
female to male mice.
For demonstration in the vignette, we have saved a small subset of
the results from this
[Zenodo entry](https://zenodo.org/records/10381745). The dataset
was made available under an MIT license. For information on how the
DTU table was saved, this is noted in `inst/scripts/make-data.R`.
Generating the exons BED file is also described there, which was
downloaded and parsed from the GENCODE M31 GTF file
(comprehensive gene annotation).
See below for more details on how to prepare exons for use with
_splicelogic_.

### Input format

The exons are as described in [input data](#input-data), and this time
the coefficient column is the one reported by the upstream DTU
method — here the *DEXSeq* `estimate`, alongside `padj` so that we can
restrict the input to significant transcripts. Its sign is what sets
the direction of the comparison, so nothing has to be partitioned by
hand.

### Loading example data

In this section we load the small example dataset that has been
prepared for this vignette. Note that in a typical *splicelogic*
workflow, the exons *GRanges* would be loaded from a *TxDb*.
See the [obtaining exon ranges](#obtaining-exon-ranges) section
below for further details.

We load the differential transcript usage analysis from the Jones
*et al.* paper. Specifically, we load the transcripts found to
exhibit DTU in the
[F vs M comparison in cortex](https://github.com/lasseignelab/230227_EJ_MouseBrainIsoDiv/blob/main/results/dtu_transcripts/cortex_sex_sig_features.csv).

The DTU table for this example, and the exons for the differential
transcripts have been saved in the `extdata` directory of
*splicelogic*.

The DTU table was generated by the authors of Jones *et al.*,
which used _IsoformSwitchAnalyzeR_ to process DTU analysis results
from _DEXSeq_, running on transcript counts instead of exon counts.
For _splicelogic_, we only require that some statistical method was
used to test DTU, that we have adjusted p-values allowing us to
subset to an FDR bounded set, and that we have a numeric value
indicating the direction of change associated with each transcript.

```{r dtu-table}
#| message: false
library(readr)
# load DTU results
dir <- system.file("extdata", package="splicelogic")
dtu_table <- readr::read_delim(file.path(dir, "dtu_table.tsv"))
```

We load from BED file the 601 exons from GENCODE M31 annotation
for the 49 differential transcripts, and attach the metadata
columns.

```{r load-exons}
#| message: false
library(plyranges)
exons_file <- "exons_M31.bed.gz"
exons_mcols_file <- "exons_mcols.tsv.gz"
exons <- plyranges::read_bed(file.path(dir, exons_file))
mcols(exons) <- DataFrame(
  readr::read_delim(file.path(dir, exons_mcols_file))
)
```

Finally, we add `Seqinfo` for mm49, the reference mouse genome for
GENCODE M31.

```{r add-seqinfo}
si <- Seqinfo::Seqinfo(genome="mm39")
seqlevels(exons) <- seqlevels(si)
seqinfo(exons) <- si
```

Next, we join the DTU results onto the `exons` GRanges. This is
necessary for the *splicelogic* functions to know which transcripts
are up- or down-regulated, and to which gene the exons and
transcripts belong.

The code below matches the transcript IDs in the `exons` metadata
with the DTU results table, and then binds the two tables before
assigning back to `exons`.

```{r merge-dtu}
txp_idx <- match(exons$tx_id, dtu_table$tx_id)
cols_to_add <- dtu_table[txp_idx, ] |> dplyr::select(-tx_id)
merged_DF <- S4Vectors::cbind(mcols(exons), cols_to_add)
mcols(exons) <- merged_DF
```

As of Bioc 3.23 and *plyranges* version 1.32, the above code to
add metadata columns from an additional table can be replaced with
the following:

```{r join-mcols}
#| eval: false
exons <- exons |>
  plyranges::join_mcols_left(dtu_table, by = "tx_id")
```

*splicelogic* is designed to take as input the exons from
significantly changed transcripts, so we first filter out
transcripts that were not signficant at FDR 10%.

```{r filter-sig}
sig_exons <- exons |> filter(padj < .1)
sig_exons |> head(2)
```

### Preprocessing input data {#preprocessing}

The first step is to run `preprocess()`, which prepares the `exons`
for event detection. This function checks the input data, ensures
that the necessary columns are present:

```{r preprocess}
#| message: false
sig_exons <- sig_exons |>
  preprocess(coef_col = "estimate")
sig_exons |> head(2)
```

`preprocess()` also accepts an optional `additional_columns`
argument. This lets us specify which metadata columns we want
to bring over from the paired transcript (`event_tx_id`) into
the event output. It takes a character vector of column names
already present in `mcols(exons)`; each will then appear in the
output as `event_<col>`. See `?preprocess` for details.

### Finding individual events {#individual-events}

Next we can run the various functions for calculating different
types of splicing events, listed in
[the `find_*()` functions](#find-functions).

**Skipped exons (SE)**

For example, we can calculate exons that are skipped in
up-regulated transcripts relative to down-regulated transcripts,
across all genes. As these are skipped in the up-regulated
transcripts, it is expected that the exons returned belong to
down-regulated transcripts:

```{r calc-se}
skipped <- sig_exons |> find_se()
skipped
```

**Included exons (IE)**

```{r calc-re}
included <- sig_exons |> find_ie()
included
```

**Mutually exclusive exons (MXE)**

```{r calc-mx}
mx <- sig_exons |> find_mxe()
mx
```

**Retained introns (RI)**

Here we do not find retained introns, and the function returns an
empty vector.

```{r calc-ri}
ri <- sig_exons |> find_ri()
ri
```

**Alternative 5' and 3' splice sites (A5/3SS)**

```{r calc-a5-a3-ss}
a5ss <- sig_exons |> find_a5ss()
a5ss
a3ss <- sig_exons |> find_a3ss()
a3ss
```

**Alternative transcription start and end sites (ATSS/ATES)**

These two compare only the first exon, or only the last exon, of each
transcript, and ask whether transcription begins or ends at a different
coordinate. Unlike the splice-site finders they do not require the two
exons to overlap, so an alternative promoter somewhere else in the gene
is still reported:

```{r calc-atss-ates}
atss <- sig_exons |> find_atss()
atss
ates <- sig_exons |> find_ates()
ates
```

### Finding all events

The following function wraps the above ones to find all events.

```{r calc-all}
all_events <- sig_exons |> find_all_events()
all_events
```

```{r events-barplot}
#| fig.alt: 'Barplot of event types'
barplot(
  table(all_events$event_type), horiz=TRUE, las=1,
  xlab="exons participating in an event"
)
```
## Upstream DTU methods {#upstream-methods}

In this section we talk about one of the two routes into
*splicelogic*: the one where the two groups of transcripts come out
of a differential transcript usage (DTU) analysis. DTU methods test whether the
relative proportions of transcripts within a gene differ between
experimental conditions. 

Any upstream DTU method that produces transcript-resolved
differential usage statistics can be used with *splicelogic*,
provided that results include:

1.  a per-transcript directional effect estimate (e.g. a model
    coefficient, change in isoform fraction, deltaPSI, etc.), and\
2.  an adjusted p-value (or equivalent significance metric by
    thresholding).

Common upstream DTU methods include:

-   [satuRn](https://bioconductor.org/packages/satuRn) — fits
    quasi-binomial generalized linear models to transcript usage
    proportions and performs scalable transcript-level DTU testing.
    Particularly well suited to larger datasets. Gene-level testing
    recommended via *stageR*.

-   [DRIMSeq](https://bioconductor.org/packages/DRIMSeq) — models
    transcript proportions within genes using a
    Dirichlet-Multinomial framework, with both gene-level and
    transcript-level testing.

-   [BANDITS](https://bioconductor.org/packages/BANDITS) — a
    Bayesian hierarchical DTU method that models transcript usage
    using a Dirichlet-Multinomial and explicitly accounts for
    mapping uncertainty using equivalence classes, with both
    gene-level and transcript-level testing.

-   [DEXSeq](https://bioconductor.org/packages/DEXSeq) — primarily
    a differential exon usage (DEU) method based on negative
    binomial GLMs, but commonly used in transcript-level DTU
    workflows (e.g. the *rnaseqDTU* Bioconductor 2018 workflow).
    The log2 fold change coefficients from *DEXSeq* can be used
    directly with *splicelogic*.

-   [limma](https://bioconductor.org/packages/limma) and
    [edgeR](https://bioconductor.org/packages/edgeR)
    (`diffSplice` / `diffSpliceDGE`) — can be used for DTU
    analyses (as well as exon-level), with both gene-level and
    transcript-level testing.

Regardless of which method is used, the per-transcript DTU
statistics (effect estimates and adjusted p-values) have to be
mapped onto the individual exons of each transcript to produce an
exon-level *GRanges*
(see [Obtaining exon ranges](#obtaining-exon-ranges)). This
annotated *GRanges* is the starting point for *splicelogic*,
beginning with `preprocess()` and then followed by event-specific
functions (`find_se()`, `find_ie()`, etc., see
[the `find_*()` functions](#find-functions)).

# Collapsing redundant events {#collapsing-events}

This section is a post-processing recipe rather than a detection step.
It applies to the output of any `find_*()` call, whether the input came
from transcript sets or from a DTU table, so it is kept separate from
the walkthroughs above. The examples reuse `mock_events` from
[the synthetic set](#transcript-sets), whose structure is small enough
to read at a glance.

Because an exon can participate in an event against more than one
partner transcript, the same range can appear on several rows. To get one 
call per exon and event type, with the partners concatenated, we can
group by `exon_id` and `event_type`, then use  *plyranges*
to collapse the ranges and concatenate the partner transcript IDs.
The new columns after grouping tell us the number of pairs (`n`) 
and collapse other informative variables.

```{r mock-collapse}
# define a helper function
my_paste <- \(x) paste(unique(x), collapse = ",")
# `exon_id` is usually found in the metadata columns in the input exons. 
# If not available, we can build it from the gene and the exon coordinates — the same 
# exon then gets the same id in every transcript that contains it:
#   mock_events <- mock_events |>
#     dplyr::mutate(exon_id = paste0(gene_id, ":", start, "-", end))
mock_events |>
  dplyr::group_by(exon_id, event_type) |>
  plyranges::reduce_ranges_directed(
    n = plyranges::n(),
    tx_id = my_paste(tx_id),
    exon_rank = my_paste(exon_rank),
    event_tx_id = my_paste(event_tx_id),
  )
```

# Identifying complex events {#complex-events}

This section is also a post-processing recipe rather than a detection
step, and likewise applies to the output of any `find_*()` call,
whatever the input was. Where the previous section grouped by exon to
answer "how many distinct events are there?", grouping by transcript
pair answers a different question: which differences co-occur in a
single isoform switch. A pair often differs in more than one way at
once — an exon skipped *and* a splice site shifted — and every event row
names both transcripts involved, so the pair is enough to recover them.
The examples again reuse `mock_events` from
[the synthetic set](#transcript-sets).

We want to identify events that happen in the same transcript pair that 
lead into complex splicicing events. To do this, we can group the events
by transcript pairs, since each event is defined by a pair of transcripts.

```{r mock-pairs}
events_by_pair <- mock_events |>
  dplyr::mutate(
    tx_pair = paste0(
      pmin(tx_id, event_tx_id), "-", pmax(tx_id, event_tx_id)
    )
  )
S4Vectors::split(events_by_pair, events_by_pair$tx_pair)
```

`tx_0-tx_5` carries two events, the skipped exon and the alternative
3' splice site. Pulling the two transcripts of a pair back out of the
preprocessed exons lets us look at the event in context:

```{r mock-pair-plot}
#| warning: false
#| fig.height: 2
#| fig.alt: 'Exon structures of the transcript pair tx_0 and tx_5'
ev <- events_by_pair |>
  dplyr::filter(tx_pair == "tx_0-tx_5") |>
  head(1)
pair <- mock |>
  dplyr::filter(tx_id %in% c(ev$tx_id, ev$event_tx_id))
wiggleplotr::plotTranscripts(
  S4Vectors::split(pair, pair$tx_id),
  rescale_introns = FALSE
)
```



# Obtaining exon ranges {#obtaining-exon-ranges}

## Using `prepare_exons()`

`prepare_exons()` extracts exon ranges from a *TxDb* object and
merges them with your DTU results table. It returns a flat
*GRanges* ready for `preprocess()` and the `find_*` functions.
Typical usage shown in the following code:

```{r prepare-demo}
#| eval: false
exons <- prepare_exons(
  txdb = <A TxDb OBJECT>,
  dtu_table = <DTU_TABLE>,
  coef_col = "estimate"
)
```

We demonstrate using GENCODE v32 (human genes).
The first step is to load a *TxDb* object. Typically, a user would
supply their own GTF to `txdbmaker::makeTxDbFromGFF()` to generate
this. For this demonstration we will load a pre-constructed *TxDb*
from Bioconductor's *AnnotationHub*.

```{r ahub}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(AnnotationDbi)
  library(GenomicFeatures)
})
ah <- AnnotationHub()
txdb <- ah[["AH75191"]] # GENCODE v32 (human)
```

```{r get-txps}
suppressPackageStartupMessages({
  library(tibble)
})
txps <- txdb |>
  AnnotationDbi::select(
    keys(txdb, "TXID"), c("TXNAME","GENEID"), "TXID"
  ) |>
  tibble::as_tibble() |>
  dplyr::select(tx_num = TXID, tx_id = TXNAME, gene_id = GENEID) |>
  dplyr::filter(!is.na(gene_id))
txps
```

```{r sim-dtu}
# simulate DTU results
sim_dtu_table <- txps |>
  dplyr::mutate(
    padj = runif(dplyr::n()),
    effect_est = rnorm(dplyr::n())
  )
sim_dtu_table
```

We now build the exons *GRanges* with `prepare_exons()`:

```{r prepare-exons}
human_exons <- prepare_exons(
  txdb, sim_dtu_table, coef_col = "effect_est", verbose = TRUE
)
human_exons <- human_exons |>
  filter(padj < .01) |>
  preprocess(coef_col = "effect_est")
```


## Manual construction

This section walks through the steps that `prepare_exons()`
performs internally. This is useful if one needs more control over
the process or want to understand how exon ranges are built from
a *TxDb*.

The following extracts the exons grouped by transcript from the
*TxDb*:

```{r extract-exons}
# extract exons as a GRangesList
exons_list <- GenomicFeatures::exonsBy(
  txdb,
  by="tx"
  )
# Our DTU table aligns with txps, which aligns with the names
# of the GRangesList. prepare_exons() handles alignment checks.
names(exons_list) <- sim_dtu_table$tx_id
```

Next, we flatten the exons:

```{r flatten-exons}
flat_exons <- unlist(exons_list)
# swap tx_id with exon_name as the names of the GRanges
flat_exons$tx_id <- names(flat_exons) # store transcript ids
names(flat_exons) <- flat_exons$exon_name
```

Finally, we add the DTU results and gene ID:

```{r add-dtu}
txp_idx <- match(flat_exons$tx_id, sim_dtu_table$tx_id)
cols_to_add <- sim_dtu_table[txp_idx,] |>
  dplyr::select(-c(tx_id, tx_num))
merged_DF <- cbind(mcols(flat_exons), cols_to_add)
mcols(flat_exons) <- merged_DF
```

# Session info

```{r session-info}
sessionInfo()
```
