---
title: "splicelogic: differential transcripts to splice events"
author:
    - name: Beatriz Campillo
      orcid: 0000-0001-7323-9125
    - name: Michael I. Love
      orcid: 0000-0001-8401-0545
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: rmarkdown::html_document
vignette: |
  %\VignetteIndexEntry{Introduction to `splicelogic`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

_splicelogic_ is an R/Bioconductor package for detecting alternative splicing events from exon-level data stored as _GRanges_ objects. Given a set of exons annotated with a coefficient column indicating differential transcript usage (DTU), _splicelogic_ identifies the following types of splicing events:

-   **Skipped exons (SE)** -- exons skipped in up-regulated transcripts
-   **Included exons (IE)** -- exons included in up-regulated transcripts
-   **Mutually exclusive exons (MXE)** -- pairs of exons where one is included and another is excluded in up-regulated transcripts compared to down-regulated transcripts
-   **Retained introns (RI)** -- intronic regions that are retained as part of an exon in up-regulated transcripts
-   **Alternative 5' (A5SS)** -- exons in up-regulated transcripts that share the same 3' splice site but differ at the 5' splice site from exons in down-regulated transcripts 
-   **Alternative 3' (A3SS)** -- as above, but differing in the 3' splice site

These are detected using a series of functions, e.g. `find_se()`, `find_ie()`, etc. described below.

# Quick start

With DTU results attached to a *GRanges* of the exons from significant transcripts, one can use the following code to identify splice events:

```{r quick-start}
#| eval: false
exons <- prepare_exons(
  txdb = <A TxDB OBJECT>,
  dtu_table = <DTU_TABLE>,
  coef_col = "estimate"
)

exons <- preprocess(exons, coef_col = "estimates")

# find skipped exons:
skipped <- exons |> find_se()
```

# Input data

_splicelogic_ assumes the user has run a differential transcript usage (DTU) or differential splicing analysis providing an error bound (e.g. adjusted p-value / FDR) and and effect estimate with direction (e.g. GLM estimated coefficient or deltaPSI) (see [upstream methods](#upstream-methods)).

It also assumes the user has obtained genomic ranges representing the exon structure of each transcript being analyzed (see [obtaining exon ranges](#obtaining-exon-ranges)).

The `exons` should be provided in a flat *GRanges* object (one range per exon), containing exon-level metadata in `mcols(exons)` such as the gene ID, transcript ID, rank in the transcript, and the estimated change or at least direction of change from the differential analysis.

Required columns for exons:

| Column | Description |
|----|----|
| `gene_id` | Gene identifier |
| `tx_id` | Transcript identifier |
| `exon_rank` | Exon rank within the transcript |
| `coef_col = "<user supplied>"` | differential effect estimate for the transcript |

The differential effect estimate column is specified by the user in `preprocess()`. This should be the differential effect estimate associated with the specific transcript containing the exon, and minimally could indicate the direction of effect with `+1/-1`. Positive values indicate up-regulated exons and negative values indicate down-regulated exons. All exons from the same transcript will share the same value for this column.

## Jones et al mouse long read dataset

For demonstration, we will use a published mouse long read dataset and its reported splicing changes.

The citation is:

> Emma F. Jones, Timothy C. Howton, Victoria L. Flanary, Amanda D. Clark, Brittany N. Lasseigne Long-read RNA sequencing identifies region- and sex-specific C57BL/6J mouse brain mRNA isoform expression and usage **Mol Brain** 17, 40 (2024). [doi: 10.1186/s13041-024-01112-7](https://doi.org/10.1186/s13041-024-01112-7)

Information about the paper, including code and publicly available data can be found at this URL:

<https://github.com/lasseignelab/230227_EJ_MouseBrainIsoDiv>

In the abstract, Jones *et al.* describe the experiment:

> To assess differences in AS across the cerebellum, cortex, hippocampus, and striatum by sex, we generated and analyzed Oxford Nanopore Technologies (ONT) long-read RNA sequencing (lrRNA-Seq) C57BL/6J mouse brain cDNA libraries. From \> 85 million reads that passed quality control metrics, we calculated differential gene expression (DGE), differential transcript expression (DTE), and differential transcript usage (DTU) across brain regions and by sex.

Here we will focus on the cortex-specific sex comparison, comparing female to male mice.

## Loading example data

In this section we load the small example dataset that has been prepared for this vignette. Note that in a typical _splicelogic_ workflow, the exons _GRanges_ would be loaded from a *TxDb*. See the [obtaining exon ranges](#obtaining-exon-ranges) section below for further details.

We load the differential transcript usage analysis from the Jones *et al.* paper. Specifically, we load the transcripts found to exhibit DTU in the 
[F vs M comparison in cortex](https://github.com/lasseignelab/230227_EJ_MouseBrainIsoDiv/blob/main/results/dtu_transcripts/cortex_sex_sig_features.csv).

The DTU table for this example, and the exons for the differential transcripts have been saved in the `extdata` directory of _splicelogic_.

```{r dtu-table}
#| message: false
library(readr)
# load DTU results 
dir <- system.file("extdata", package="splicelogic")
dtu_table <- readr::read_delim(file.path(dir, "dtu_table.tsv"))
```

We load from BED file the 601 exons from GENCODE M31 annotation for the 49 differential transcripts, and attach the metadata columns.

```{r load-exons}
#| message: false
library(plyranges)
exons_file <- "exons_M31.bed.gz"
exons_mcols_file <- "exons_mcols.tsv.gz"
exons <- plyranges::read_bed(file.path(dir, exons_file))
mcols(exons) <- DataFrame(readr::read_delim(file.path(dir, exons_mcols_file)))
```

Finally, we add `Seqinfo` for mm49, the reference mouse genome for GENCODE M31.

```{r add-seqinfo}
si <- Seqinfo::Seqinfo(genome="mm39")
seqlevels(exons) <- seqlevels(si)
seqinfo(exons) <- si
```

Next, we join the DTU results onto the `exons` GRanges. This is necessary for the _splicelogic_ functions to know which transcripts are up- or down-regulated, and to which gene the exons and transcripts belong. 

The code below matches the transcript IDs in the `exons` metadata with the DTU results table, and then binds the two tables before assigning back to `exons`.

```{r merge-dtu}
txp_idx <- match(exons$tx_id, dtu_table$tx_id)
cols_to_add <- dtu_table[txp_idx, ] |> dplyr::select(-tx_id)
merged_DF <- S4Vectors::cbind(mcols(exons), cols_to_add)
mcols(exons) <- merged_DF
```

As of Bioc 3.23 and *plyranges* version 1.32, the above code to add metadata columns from an additional table can be replaced with the following:

```{r join-mcols}
#| eval: false
exons <- exons |> 
  plyranges::join_mcols_left(dtu_table, by = "tx_id")
```

_splicelogic_ is designed to take as input the exons from significantly changed transcripts, so we first filter out transcripts that were not signficant at FDR 10%.

```{r filter-sig}
sig_exons <- exons |> filter(padj < .1)
```

# Finding splicing events

## Preprocessing input data

The first step is to run `preprocess()`, which prepares the `exons` for event detection. This function checks the input data, ensures that the necessary columns are present:

```{r preprocess}
#| message: false
library(splicelogic)
sig_exons <- sig_exons |> 
  preprocess(coef_col = "estimate")
```

## Finding individual events

Next we can run the various functions for calculating different types of splicing events. 

**Skipped exons (SE)**

For example, we can calculate exons that are skipped in up-regulated transcripts relative to down-regulated transcripts, across all genes. As these are skipped in the up-regulated transcripts, it is expected that the exons returned belong to down-regulated transcripts:

```{r calc-se}
skipped <- sig_exons |> find_se()
skipped
```

**Included exons (IE)**

```{r calc-re}
included <- sig_exons |> find_ie()
included
```

**Mutually exclusive exons (MXE)**

```{r calc-mx}
mx <- sig_exons |> find_mxe()
mx
```

**Retained introns (RI)**

Here we do not find retained introns, and the function returns an empty vector.

```{r calc-ri}
ri <- sig_exons |> find_ri()
ri
```

**Alternative 5' and 3' splice sites (A5/3SS)**

```{r calc-a5-a3-ss}
a5ss <- sig_exons |> find_a5ss()
a5ss
a3ss <- sig_exons |> find_a3ss()
a3ss
```

## Finding all events

The following function wraps the above ones to find all events.

```{r calc-all}
all_events <- sig_exons |> find_all_events()
all_events
```

```{r events-barplot}
#| fig.alt: 'Barplot of event types'
barplot(
  table(all_events$event), horiz=TRUE, las=1, 
  xlab="exons participating in an event"
)
```

# Upstream methods

`splicelogic` is designed to operate downstream of **differential transcript usage (DTU)**. DTU methods test whether the **relative proportions of transcripts within a gene** differ between experimental conditions.

In general, any upstream method that produces transcript-resolved differential usage statistics can be used with `splicelogic`, provided that results include:

1. a **per-transcript directional effect estimate** (e.g. a model coefficient, change in isoform fraction, deltaPSI, etc.), and  
2. an **adjusted p-value** (or equivalent significance metric by thresholding).

Common upstream methods include:

- [satuRn](https://bioconductor.org/packages/satuRn) — fits quasi-binomial generalized linear models to transcript usage proportions and performs scalable transcript-level DTU testing. Particularly well suited to larger datasets.

- [DRIMSeq](https://bioconductor.org/packages/DRIMSeq) — models transcript proportions within genes using a Dirichlet-multinomial framework, with both gene-level and transcript-level testing.

- [BANDITS](https://bioconductor.org/packages/BANDITS) — a Bayesian hierarchical DTU method that models transcript usage with a Dirichlet-multinomial and explicitly accounts for mapping uncertainty using equivalence classes, with both gene-level and transcript-level testing.

- [DEXSeq](https://bioconductor.org/packages/DEXSeq) — primarily a differential exon usage (DEU) method based on negative binomial GLMs, but commonly used in transcript-level DTU workflows. The log2FC coefficients from DEXSeq can be used directly with _splicelogic_.

- [limma](https://bioconductor.org/packages/limma) and [edgeR](https://bioconductor.org/packages/edgeR) (`diffSplice` / `diffSpliceDGE`) — can be used for DTU analyses (as well as exon-level), with both gene-level and transcript-level testing.

Regardless of which method is used, the per-transcript DTU statistics (effect estimates and adjusted p-values) have to be mapped onto the individual exons of each transcript to produce an exon-level _GRanges_ (see [Obtaining exon ranges](#obtaining-exon-ranges)). This annotated _GRanges_ is the starting point for _splicelogic_, beginning with `preprocess()` and then followed by event-specific functions (`find_se()`, `find_ie()`, etc.).

# Obtaining exon ranges

## Using `prepare_exons()`

`prepare_exons()` extracts exon ranges from a *TxDb* object and merges them with your DTU results table. It returns a flat *GRanges* ready for `preprocess()` and the `find_*` functions. functions. We demonstrate using GENCODE v32 (human genes).

The first step is to load a _TxDb_ object. Typically, a user would supply their own GTF to `txdbmaker::makeTxDbFromGFF()` to generate this. For this demonstration we will load a pre-constructed _TxDb_ from Bioconductor's _AnnotationHub_.

```{r ahub}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(AnnotationDbi)
  library(GenomicFeatures)
})
ah <- AnnotationHub()
txdb <- ah[["AH75191"]] # GENCODE v32 (human)
```

```{r get-txps}
suppressPackageStartupMessages({
  library(tibble)
})
txps <- txdb |>
  AnnotationDbi::select(keys(txdb, "TXID"), c("TXNAME","GENEID"), "TXID") |>
  tibble::as_tibble() |>
  dplyr::select(tx_num = TXID, tx_id = TXNAME, gene_id = GENEID)
txps
```

```{r sim-dtu}
# simulate DTU results
sim_dtu_table <- txps |>
  dplyr::mutate(
    padj = runif(dplyr::n()),
    effect_est = rnorm(dplyr::n())
  )
sim_dtu_table
```

Here we name this output `human_exons` to not collide with the example above for the mouse dataset.

```{r prepare-exons}
human_exons <- prepare_exons(txdb, sim_dtu_table, coef_col = "effect_est", verbose = TRUE)
human_exons <- human_exons |>
  filter(padj < .01) |>
  preprocess(coef_col = "effect_est")
```

## Manual construction

This section walks through the steps that `prepare_exons()` performs internally. This is useful if you need more control over the process or want to understand how exon ranges are built from a *TxDb*.

The following extracts the exons grouped by transcript from the *TxDb*:

```{r extract-exons}
# extract exons as a GRangesList
exons_list <- GenomicFeatures::exonsBy(
  txdb,
  # filter = AnnotationFilter::TxIdFilter(unique(sim_dtu_table$tx_id)), 
  by="tx"
  )
# Our DTU table aligns with txps, which aligns with the names of the GRangesList.
# prepare_exons() handles this with alignment checks
names(exons_list) <- sim_dtu_table$tx_id
```

Next flattening the exons (here using a new name `flat_exons` to not collide with the `exons` mouse dataset above):

```{r flatten-exons}
flat_exons <- unlist(exons_list)
# swap tx_id with exon_name as the names of the GRanges
flat_exons$tx_id <- names(flat_exons) # store transcript ids
names(flat_exons) <- flat_exons$exon_name
```

Adding DTU results and gene ID:

```{r add-dtu}
txp_idx <- match(flat_exons$tx_id, sim_dtu_table$tx_id)
cols_to_add <- sim_dtu_table[txp_idx,] |> dplyr::select(-c(tx_id, tx_num))
merged_DF <- cbind(mcols(flat_exons), cols_to_add)
mcols(flat_exons) <- merged_DF
```

# Session info

```{r session-info}
sessionInfo()
```
