---
title: "Introduction to `splicelogic`"
author:
    - name: Beatriz Campillo
      orcid: 0000-0001-7323-9125
    - name: Michael I. Love
      orcid: 0000-0001-8401-0545
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: rmarkdown::html_document
vignette: |
  %\VignetteIndexEntry{Introduction to `splicelogic`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

`splicelogic` is an R/Bioconductor package for detecting alternative
splicing events from exon-level data stored as `GRanges` objects.
Given a set of exons annotated with a coefficient column indicating
differential transcript usage (positDTU), `splicelogic` identifies
the following types of splicing events:

- **Skipped exons (SE)** -- exons present in one isoform but absent
  in another.
- **Mutually exclusive exons (MXE)** -- pairs of exons where one is
  included at the expense of the other.
- **Retained introns (RI)** -- intronic regions that are retained as
  part of an exon in an alternative isoform.
- **Alternative 3' (A3SS)** -- exons that share the same 5' splice site
  but differ at the 3' end.
- **Alternative 5' (A5SS)** -- exons that share the same 3' splice site
  but differ at the 5' end.

# Input format

`splicelogic` operates on `GRanges` objects that contain exon-level
annotations. Each range represents one exon and must carry the
following metadata columns:

| Column      | Description                                        |
|-------------|----------------------------------------------------|
| `gene_id`   | Gene identifier                                    |
| `tx_id`     | Transcript identifier                              |
| `exon_rank` | Exon rank within the transcript                     |
| *coef_col*  | DTU coefficient (column name is user-defined) 

<!--
considering if exon_rank should be inputed by user or calculated internally
currently, it is required as input
-->


The coefficient input in *coef_col* indicates the differential transcript usage (DTU) of the specific
transcript containing the exon. All exons from the same transcript share the
same coefficient value. This values comes from a prior differential transcript usage
analysis, such as that performed by `satuRn`.
Positive coefficient values indicate upregulated exons and negative
values indicate downregulated exons. 

# How to start detecting splicing events

```{r setup, message = FALSE}
# library(splicelogic)
devtools::load_all()
```


## Preprocessing
We can create mock data for demonstration purposes using the
`create_mock_data()` function. Before running any event detection, 
call `preprocess_input()` to
validate the data and annotate each exon with internal flags and a
unique key:

```{r preprocess}
gr <- create_mock_data( n_genes = 2, n_tx = 4, n_exons = 6 )
gr
gr <- preprocess_input(gr, coef_col = "coefs")
mcols(gr)
```

## Skipped exons

`calc_skipped_exons()` identifies exons that are present in
downregulated transcripts but absent (skipped) in upregulated ones.

We use the `se_mock_data()` dataset, which contains one gene with
three transcripts: one downregulated (7 exons) and two upregulated
(6 exons each, missing exons at ranks 3 and 5 of the reference
transcript).

```{r skipped-exons}
gr_se <- se_mock_data()
gr_se <- preprocess_input(gr_se, coef_col = "coefs")
se_result <- calc_skipped_exons(gr_se, coef_col = "coefs")
se_result
```

The result is a `GRanges` object containing only the exons flagged as
skipped, with an `event` column set to `"skipped_exon"`.

```{r skipped-exons-check}
mcols(se_result)[, c("gene_id", "tx_id", "exon_rank", "event")]
```

## Mutually exclusive exons

`calc_mutually_exclusive()` detects pairs of exons where one exon is
included in one isoform and a different exon occupies the equivalent
position in another isoform.

The `mx_mock_data()` dataset is designed to contain mutually exclusive
exon events:

```{r mx-exons}
gr_mx <- mx_mock_data()
gr_mx <- preprocess_input(gr_mx, coef_col = "coefs")
mx_result <- calc_mutually_exclusive(gr_mx, coef_col = "coefs")
mx_result
```

```{r mx-exons-check}
mcols(mx_result)[, c("gene_id", "tx_id", "exon_rank", "event")]
```

## Retained introns

`calc_retained_introns()` finds exons in upregulated transcripts that
overlap intronic regions of downregulated transcripts, indicating
intron retention.

```{r retained-introns}
gr_ri <- create_mock_data(3, 6, 3)
gr_ri <- preprocess_input(gr_ri, coef_col = "coefs")
gr_ri <- generate_retained_introns(gr_ri, n_ri = 3)
ri_result <- calc_retained_introns(gr_ri, coef_col = "coefs")
ri_result
```

```{r retained-introns-check}
if (length(ri_result) > 0) {
    mcols(ri_result)[, c("gene_id", "tx_id", "exon_rank", "event")]
}
```

## Alternative 3' and 5' splice sites

`calc_a3ss_a5ss()` detects exons that share one boundary (start or
end) with an exon in another isoform but differ at the opposite end,
indicating alternative splice site usage.

```{r a3ss-a5ss}
gr_ss <- create_mock_data(3, 3, 6)
gr_ss <- preprocess_input(gr_ss, coef_col = "coefs")
gr_ss <- generate_a3ss(gr_ss, n_a3ss = 3)
ss_result <- calc_a3ss_a5ss(gr_ss, coef_col = "coefs")
ss_result
```

```{r a3ss-a5ss-check}
if (length(ss_result) > 0) {
    mcols(ss_result)[, c("gene_id", "tx_id", "exon_rank", "event")]
}
```

# Handling cases with no events

When the input data does not contain any splicing events, each
detection function returns an empty `GRanges` object:

```{r no-events}
gr_none <- no_event_mock_data()
gr_none <- preprocess_input(gr_none, coef_col = "coefs")

calc_skipped_exons(gr_none, coef_col = "coefs")
calc_mutually_exclusive(gr_none, coef_col = "coefs")
```

# Session info

```{r session-info}
sessionInfo()
```
